---
title: "Census Tract/Block Query per County"
author: "Gabriel"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Load neccesary libraries 

install.packages('tidyverse')
install.packages('tidycensus')
install.packages('sf')
install.packages('kableExtra')
library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")


#Load API KEY
census_api_key("3aedcfd189958d66a813a29a4b33580e78657644", overwrite = TRUE)
#Set CRS
crs<-4326



```

USE LOAD VARIABLES TO EXPLORE DATA OF INTEREST
```{r}
acs_variable_list.2009 <- load_variables(2009, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)
#USe filter to filter by some title in  the interphase
#list of variables and the rename function
#add the E to get the estimate, otherwise returns margin as well
acs_vars <- c("B06003_013E", #Foreign Born - PLACE OF BIRTH BY SEX IN THE UNITED STATES
              "B05002_013E", #Estimate!!Total!!Foreign born - PLACE OF BIRTH BY CITIZENSHIP STATUS
              "B05007_039E", #Estimate!!Total!!Latin America!!Caribbean!!- Entered before 1980
              "B25002_003E", #stimate!!Total!!Vacant OCCUPANCY STATUS
              "B09016_002E", #Estimate!!Total!!In households - HOUSEHOLD TYPE (INCLUDING LIVING ALONE) BY RELATIONSHIP
              "B19013_001E", #Estimate!!Median household income in the past 12 months (in 2009 inflation-adjusted dollars)
              "B19037A_053E", #Estimate!!Total!!Householder 65 years and over
              "B06012_002E", #Total!!Below 100 percent of the poverty level
              "B25071_001E", #Median gross rent as a percentage of household income
              "B25035_001E", #Median year structure built
              "B19057_001E", #Total - PUBLIC ASSISTANCE INCOME IN THE PAST 12 MONTHS FOR HOUSEHOLDS
              "B25001_001E", #Total HOUSING UNITS
              "B05006_127E", #Total!!Americas!!Latin America!!Caribbean!!Cuba
              "B25077_001E") #med values


```

Interesting Tables:
ACS 2009 5-Year Estimates
B06003_013 - Foreign Born - PLACE OF BIRTH BY SEX IN THE UNITED STATES
B05002_013 - Estimate!!Total!!Foreign born - PLACE OF BIRTH BY CITIZENSHIP STATUS
B05007_039 - Estimate!!Total!!Latin America!!Caribbean!!- Entered before 1980
B25002_003 - Estimate!!Total!!Vacant OCCUPANCY STATUS
B09016_002 - Estimate!!Total!!In households - HOUSEHOLD TYPE (INCLUDING LIVING ALONE) BY RELATIONSHIP
B19013_001 - Estimate!!Median household income in the past 12 months (in 2009 inflation-adjusted dollars) MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS (IN 2009 INFLATION-ADJUSTED DOLLARS)
B19037A_053 - Estimate!!Total!!Householder 65 years and over
B06012_002 - Estimate!!Total!!Below 100 percent of the poverty level
B25071_001 - Estimate!!Median gross rent as a percentage of household income
B25035_001 - Estimate!!Median year structure built
B19057_001 - Estimate!!Total - PUBLIC ASSISTANCE INCOME IN THE PAST 12 MONTHS FOR HOUSEHOLDS
B25001_001 - Estimate!!Total HOUSING UNITS
B05006_127 - Estimate!!Total!!Americas!!Latin America!!Caribbean!!Cuba

```{r}
#The Census Query sample
jackson_county_tracts.09 <- get_acs(geography = "tract",
                             year = 2009, 
                             variables = acs_vars, 
                             geometry = TRUE, 
                             state = "MS", 
                             county = "Hinds", 
                             output = "wide") 

#Rename the variables

jackson_county_tracts.09 <- jackson_county_tracts.09%>%
  rename(fb_placebysex.2009 = B06003_013E,
          totalfb_placebycitizenshipstatus.2009 = B05002_013E,
          total_carib_1980.2009 = B05007_039E,
          total_vacancy_O_status.2009 = B25002_003E,
          total_householdtype_relationship.09 = B09016_002E,
          med_hh_inc.09 = B19013_001E,
          hh_65up.09 = B19037A_053E,
          Total_b_100pov.09 = B06012_002E,
          gross_v_income_percentage.09 = B25071_001E,
          med_built_year.09 = B25035_001E,
          tot_publicass_inc.09 = B19057_001E,
          tot_housingunits.09 = B25001_001E,
          cub_fb_total09 = B05006_127E,
          med_houseval09 = B25077_001E
          )%>%
  mutate(vacancyPct.2009 = total_vacancy_O_status.2009/tot_housingunits.09) %>% # Get Vacanct Rate
   st_as_sf(crs = crs)
#setcrs
```
2022 ACS 5 year Estimates

```{r}

#Load Variables
acs_variable_list.22 <- load_variables(2022, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

acs22_vars <-c("B06012PR_002E", #Tot Below 100 percent of the poverty level
               "B25002_003E", #TotalVacant
               "B21004_001E", #Median income in the past 12 months (in 2022 inflation-adjusted dollars
               "B09021_023E", #65 years and over Lives alone
               "B25136_001E", #Total Housing Units
               "B25071_001E", #Median gross rent as a percentage of household income
               "B05006_143E", #Total:!!Americas:!!Latin America:!!Caribbean:!!Cuba
               "B19057_002E", #Total:!!With public assistance income
               "B25035_001E", #Median year structure built
               "B25077_001E" #med values
       )
#Going to get the same data for 15 years later
jackson_county_tracts.22 <- get_acs(geography = "tract",
                             year = 2022, 
                             variables = acs22_vars, 
                             geometry = TRUE, 
                             state = "MS", 
                             county = "Hinds", 
                             output = "wide") 

jackson_county_tracts.22 <- jackson_county_tracts.22%>%
  rename(tot_below100pov22 = B06012PR_002E,
         tot_vacant22 = B25002_003E,
         med_inc22 = B21004_001E,
         up65_alone22 = B09021_023E,
         tot_housingunits22 =B25136_001E,
         gross_v_inc_perc22 = B25071_001E,
         tot_fb_cuba22 = B05006_143E,
         tot_pa_inc22 = B19057_002E,
         med_yb22 = B25035_001E,
         med_values22 = B25077_001E)%>%
  mutate(vacpct22 = (tot_vacant22/tot_housingunits22)) %>%
  st_as_sf(crs = crs)
```

Interesting Variables:

2022 5 Year Estimates

B06012PR_002 - Estimate!!Total:!!Below 100 percent of the poverty level
B25002_003 - Estimate!!Total:!!Vacant
B21004_001 - Estimate!!Median income in the past 12 months (in 2022 inflation-adjusted dollars)--!!Total
B09021_023 - Estimate!!Total:!!65 years and over:!!Lives alone
B25136_001 - Estimate!!Total Housing Units:
B25071_001 - Estimate!!Median gross rent as a percentage of household income
B05006_143 - Estimate!!Total:!!Americas:!!Latin America:!!Caribbean:!!Cuba
B19057_002 - Estimate!!Total:!!With public assistance income
B25035_001 - Estimate!!Median year structure built

```{r}
#Join the two tables to get an entire dataset to work with
allACS_df <- st_drop_geometry(jackson_county_tracts.22,jackson_county_tracts.09)%>%
  left_join(jackson_county_tracts.09 , jackson_county_tracts.22,
                    by= c("GEOID"))
#create variables of changes

allACS_df <- allACS_df %>%
  mutate(change_vac_pct = vacpct22 - vacancyPct.2009,
         change_med_inc = med_inc22 - med_hh_inc.09,
         change_med_values = med_values22 - med_houseval09,
         change_cuba_fb = tot_fb_cuba22 - cub_fb_total09,
         change_pct_below100pov = tot_below100pov22 - Total_b_100pov.09)

sf.allACS <- st_join(jackson_county_tracts.09, jackson_county_tracts.22, left= TRUE ) %>%
    mutate(change_vac_pct = vacpct22 - vacancyPct.2009,
         change_med_inc = med_inc22 - med_hh_inc.09,
         change_med_values = med_values22 - med_houseval09,
         change_cuba_fb = tot_fb_cuba22 - cub_fb_total09,
         change_pct_below100pov = tot_below100pov22 - Total_b_100pov.09)
```
You can now create a list of census tracts within the chosen county and use it to creat a label that can the be used to group by and summarize how those areas compare to the rest of the city. 

1. create vector of census tract geoids
```{r}
# Make plots of the data
ggplot()+
  geom_sf(data = sf.allACS, aes(fill = change_med_inc),
          color = "transparent")
 

```

Bring in features to join to the census data. In this case, separate dataframes of stops for each train line in the form of points for each row and are stacked together using rbind. In the example below, a label is created to distinguish which set of points belonged to what line within the aggregated dataframe.

```{r}
#example using train stop points

el <- st_read("https://opendata.arcgis.com/datasets/8c6e2575c8ad46eb887e6bb35825e1a6_0.geojson")
Broad_St <- st_read("https://opendata.arcgis.com/datasets/2e9037fd5bef406488ffe5bb67d21312_0.geojson")

septaStops <- 
  rbind(
     el %>% 
      mutate(Line = "El") %>%
      dplyr::select(Station, Line),
     Broad_St %>%
      mutate(Line ="Broad_St") %>%
      dplyr::select(Station, Line)) %>%
  st_transform(st_crs(tracts16))  #sets crs to match that of another sf object (tract 16's)

#Clip operation

clip <- 
  st_intersection(buffer, tracts16) %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Clip")

# Spatial intersection with `st_intersects()` on polygons
#     Do a spatial selection to see which tracts touch the buffer

#     approach #1: sub-setting a spatial object with a spatial object using the '[' brackets.

selection1 <- 
  tracts16[buffer,] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

#In this usage, the intersecting rows of `tracts16` are selected for by indexing on `buffer`. This method is very fast to compute, but hard to understand at first.

#Another Approach

# Finally, we are using centroids, or the point at the center of a polygon, to select whether a polygon from `tracts16` intersects the `buffer`.
# first they create the buffered features and drop the geometry. it is a spatial operation. to get the data of variables they merge to the original data (tract 16query) based on GEOID of the clipped list of tracts

selectCentroids <-
  st_centroid(tracts16)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts16, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")

#1  convert tract polygons to centroids and to the spatial intersect to buffers using the `selection1` approach from above. The resulting geometry type is now a point and not a polygon

#2  drop the geometry so that is becomes a regular non-spatial data frame

#3  join the original `tracts16` data frame using the `GEOID` as the join field. Note that only the `GEOID` column is selected, however the polygon spatial features are also joined.

#4  `st_sf()` is a function to turn it back to an {sf} spatial object

#5  finally the `TotalPop` column is selected and a new field of `Selection_Type` is added

#Note: you will get a warning about "st_centroid assumes attributes... don't worry about it.

```

CREATE SUMMARY TABLES

```{r}
# Example

allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent, na.rm = T),
            Population = mean(TotalPop, na.rm = T),
            Percent_White = mean(pctWhite, na.rm = T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Percent_Poverty = mean(pctPoverty, na.rm = T))

kable(allTracts.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.2")
```



```{r}
```
