---
title: "ARSHAD REPORT"
author: "Saturnya"
date: "2025-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidycensus)
library(dplyr)
library(sf)
library(here)

options(tigris_use_cache = TRUE)
options(scipen=999)
options(tigris_class = "sf")
options(tigris_cache_dir = "cache/tigris_cache")

#crs for austin

crs<-"EPSG:2277"

tidycensus::census_api_key("811e1f6f8d1f299cb75a0f0c07e01aafd801fa79", overwrite = TRUE)

```

15 minutes-30 minutes away from home
10 minutes outside of the highway
At least 5 parking spaces on site
1 story storefront typology(check with the appraiser and the property class dictionary
Middle/high incomes
30-50 Years old
Singles
$20 per sqft price -5,000 sqft needed ($10,000 annual rent price)
Maximize distance from other medical spas


Order: 

-Building Code for Distance Accumualtion layer
-Census call for variables of interest
-Local Moran's Hotspots
-reclassify accumulation layerand assign the value of a zonal analysis to tracts. Show in map tracts that are hotspots and nearest the highway
-Overlay distance accumulation with hot spots
-Get medical Spas put on the map

04/10/2025

Need to clean up all the maps with erase water and bring in the road network and the house's location



##Project specific Census Call

###get 2023 variables

```{r}
# === Variable Vector with Inline Descriptions ===
all_vars <- c(
  # === Income ===
  "B19013_001",  # Median household income
  "B19025_001",  # Aggregate household income

  # Family income brackets for $75k+ and $125k+
  "B19101_013",  # $75k–$99k
  "B19101_014",  # $100k–$124k
  "B19101_015",  # $125k–$149k
  "B19101_016",  # $150k–$199k
  "B19101_017",  # $200k+

  # === Race / Ethnicity ===
  "C02003_003",  # White alone
  "C02003_004",  # Black alone
  "C02003_006",  # Asian alone
  "C02003_009",  # Two or more races
  "B03003_001",  # Hispanic or Latino

  # === Education ===
  "B15003_022",  # Bachelor's degree
  "B15003_023",  # Master's degree
  "B15003_024",  # Professional degree
  "B15003_025",  # Doctorate

  # === Population & Age ===
  "B01003_001",  # Total population
  "B01002_001",  # Median age (Total)
  "B01002_002",  # Median age (Male)
  "B01002_003",  # Median age (Female)

  # === Housing & Vacancy ===
  "B25107_011",  # Median value homes built before 1939
  "B25005_002",  # Vacant homes (residence elsewhere)

  # === Living Alone ===
  "B09019_005",  # Males living alone
  "B09019_008",  # Females living alone

  # === Ancestry ===
  "B05006_143",  # Cuban ancestry

  # === Male 30–54 ===
  "B12002_008", "B12002_009", "B12002_010", "B12002_011", "B12002_012",  # Never married
  "B12002_040", "B12002_041", "B12002_042", "B12002_043", "B12002_044",  # Separated
  "B12002_070", "B12002_071", "B12002_072", "B12002_073", "B12002_074",  # Widowed
  "B12002_085", "B12002_086", "B12002_087", "B12002_088", "B12002_089",  # Divorced

  # === Female 30–54 ===
  "B12002_101", "B12002_102", "B12002_103", "B12002_104", "B12002_105",  # Never married
  "B12002_133", "B12002_134", "B12002_135", "B12002_136", "B12002_137",  # Separated
  "B12002_163", "B12002_164", "B12002_165", "B12002_166", "B12002_167",  # Widowed
  "B12002_178", "B12002_179", "B12002_180", "B12002_181", "B12002_182"   # Divorced
)
```

```{r}



```
###Get and process the census tract data for Austin
make the call and rename as neccesary. Follow this up with a mutation of calculations

Ideally, use the code at the bottom to filter out unreliable variables and the nsee what we can confidently calculate.

```{r}
# === Retrieve Data ===
tract_data <- get_acs(
  geography = "tract",
  variables = all_vars,
  state = "TX",
  county = "Travis",
  year = 2023,
  geometry = TRUE,
  survey = "acs5",
  output = "wide"
)

tract_data_final <- tract_data %>%
  rename(
    median_household_income = B19013_001E,
    aggregate_household_income = B19025_001E,
    income_75k_99k = B19101_013E,
    income_100k_124k = B19101_014E,
    income_125k_149k = B19101_015E,
    income_150k_199k = B19101_016E,
    income_200k_plus = B19101_017E,
    white_alone = C02003_003E,
    black_alone = C02003_004E,
    asian_alone = C02003_006E,
    two_or_more_races = C02003_009E,
    hispanic_or_latino = B03003_001E,
    bachelors_degree = B15003_022E,
    masters_degree = B15003_023E,
    professional_degree = B15003_024E,
    doctorate = B15003_025E,
    total_population = B01003_001E,
    median_age_total = B01002_001E,
    median_age_male = B01002_002E,
    median_age_female = B01002_003E,
    median_value_pre1939 = B25107_011E,
    vacant_residence_elsewhere = B25005_002E,
    male_living_alone = B09019_005E,
    female_living_alone = B09019_008E,
    cuban_ancestry = B05006_143E,
    male_30_34_never = B12002_008E,
    male_35_39_never = B12002_009E,
    male_40_44_never = B12002_010E,
    male_45_49_never = B12002_011E,
    male_50_54_never = B12002_012E,
    male_30_34_sep = B12002_040E,
    male_35_39_sep = B12002_041E,
    male_40_44_sep = B12002_042E,
    male_45_49_sep = B12002_043E,
    male_50_54_sep = B12002_044E,
    male_30_34_wid = B12002_070E,
    male_35_39_wid = B12002_071E,
    male_40_44_wid = B12002_072E,
    male_45_49_wid = B12002_073E,
    male_50_54_wid = B12002_074E,
    male_30_34_div = B12002_085E,
    male_35_39_div = B12002_086E,
    male_40_44_div = B12002_087E,
    male_45_49_div = B12002_088E,
    male_50_54_div = B12002_089E,
    female_30_34_never = B12002_101E,
    female_35_39_never = B12002_102E,
    female_40_44_never = B12002_103E,
    female_45_49_never = B12002_104E,
    female_50_54_never = B12002_105E,
    female_30_34_sep = B12002_133E,
    female_35_39_sep = B12002_134E,
    female_40_44_sep = B12002_135E,
    female_45_49_sep = B12002_136E,
    female_50_54_sep = B12002_137E,
    female_30_34_wid = B12002_163E,
    female_35_39_wid = B12002_164E,
    female_40_44_wid = B12002_165E,
    female_45_49_wid = B12002_166E,
    female_50_54_wid = B12002_167E,
    female_30_34_div = B12002_178E,
    female_35_39_div = B12002_179E,
    female_40_44_div = B12002_180E,
    female_45_49_div = B12002_181E,
    female_50_54_div = B12002_182E
  )

```
###Add calculated variables

```{r}


# === Add Derived Metrics ===
austin_tract_df <- tract_data_final %>%
  mutate(
    # Income thresholds
    income_75k_plus = income_75k_99k + income_100k_124k + income_125k_149k + income_150k_199k + income_200k_plus,
    income_125k_plus = income_125k_149k + income_150k_199k + income_200k_plus,
    pct_income_75k_plus = income_75k_plus / aggregate_household_income * 100,
    pct_income_125k_plus = income_125k_plus / aggregate_household_income * 100,

    # Race/Ethnicity
    poc_total = black_alone + two_or_more_races + hispanic_or_latino,
    pct_white = white_alone / total_population * 100,
    pct_asian = asian_alone / total_population * 100,
    pct_poc = poc_total / total_population * 100,

    # Education
    bachelors_only = bachelors_degree,
    grad_plus = rowSums(across(c(masters_degree, professional_degree, doctorate)), na.rm = TRUE),
    pct_bachelors = bachelors_only / total_population * 100,
    pct_gradplus = grad_plus / total_population * 100,

    # Living Alone
    living_alone_total = rowSums(across(c(male_living_alone, female_living_alone)), na.rm = TRUE),
    pct_living_alone = living_alone_total / total_population * 100,

    # Adults 30–54
    male_30_54 = rowSums(across(c(
      male_30_34_never, male_35_39_never, male_40_44_never, male_45_49_never, male_50_54_never,
      male_30_34_sep, male_35_39_sep, male_40_44_sep, male_45_49_sep, male_50_54_sep,
      male_30_34_wid, male_35_39_wid, male_40_44_wid, male_45_49_wid, male_50_54_wid,
      male_30_34_div, male_35_39_div, male_40_44_div, male_45_49_div, male_50_54_div
    )), na.rm = TRUE),
    
    female_30_54 = rowSums(across(c(
      female_30_34_never, female_35_39_never, female_40_44_never, female_45_49_never, female_50_54_never,
      female_30_34_sep, female_35_39_sep, female_40_44_sep, female_45_49_sep, female_50_54_sep,
      female_30_34_wid, female_35_39_wid, female_40_44_wid, female_45_49_wid, female_50_54_wid,
      female_30_34_div, female_35_39_div, female_40_44_div, female_45_49_div, female_50_54_div
    )), na.rm = TRUE),

    pct_male_30_54 = male_30_54 / total_population * 100,
    pct_female_30_54 = female_30_54 / total_population * 100,
    combined_singles= female_30_54 + male_30_54,
    combined_percent_singles=pct_male_30_54+pct_female_30_54
  )

```
###Cache Austin Dataset for future analysis
Save the final tract data. This will be combined with the hotspot analysis results to generate a masterdata geodataframe.


```{r}
saveRDS(
  austin_tract_df,
  file = here("clients", "arshad", "cache", "austin_acs5_2023_tract.rds")
)

#to load later
austin_tract_df <- readRDS(
  here("clients", "arshad", "cache", "austin_acs5_2023_tract.rds")
)


```
##Hot Spot analysis  with local morans (prep)

NOw that we have the object. Let's select the variables for hot spot analysis and the geometry.

We will then output this object in a separate script that to create local moran's I analysis and visualizations for variables.

###create vector of variables local morans

Create the vector of variables you want to select including NAME and GEOID.

Use these to filter out from the place_tract_df. Save this new data frame, it is the input tho the project's local moran's I related script

```{r}
#create vector of the variables that you want to analyze, plus geometry

locmor_austin_vars<- c("median_household_income", 
                       "white_alone", 
                       "total_population",
                       "vacant_residence_elsewhere", 
                       "income_125k_plus", 
                       "pct_125k_plus", 
                       "pct_gradplus", 
                       "grad_plus", 
                       "pct_living_alone", 
                       "combined_singles", 
                       "combined_percent_singles", 
                       "female_30_54", 
                       "male_30_54",
                       "NAME",
                       "GEOID")

austintx_locmor_df_input <- austin_tract_df %>%
  
               select(median_household_income, white_alone, total_population, vacant_residence_elsewhere, income_125k_plus,  pct_income_125k_plus, pct_gradplus, grad_plus, pct_living_alone, combined_singles, combined_percent_singles, female_30_54, male_30_54,GEOID,NAME, geometry
  )
```

Next, Cache the local moran's I object for analysis in another script. 

```{r}
saveRDS(
  austintx_locmor_df_input,
  file = here("clients", "arshad", "cache", "austin_acs5_2023_tract_locmor_input.rds")
)



```

##Local Moran's Hotspots

We now have a austing
###Load from Local morans' stat script

There is a script in the toolkit that generates the local moran's I hot spot information. 

I used it to create a df of the statistically significant hot spots. We will join these datasets together.

First, load the local morans I output df.

```{r}
locmor_austin_houtput_df<-readRDS(here("clients","arshad","cache","locmor_austin_df_houtput.rds"))

```
### combine the local morans I with the master data frame

```{r}


```

###cache that for now


##Load Austin, TX MSA Boundary

```{r}
#loaded from previous script that generates this feature

austin_cbsa<-readRDS(here("clients","arshad","cache","austin_cbsa_boundary.rds"))


```

##Clean up cartography by removing water bodies with tigris functions
###load tigris and clip the total tracts geodataframe

```{r}
#load the census tracts for the countyy

# Load water bodies for Travis County
water_travis <- area_water(state = "TX", county = "Travis", year = 2023)

# Ensure both are in the same coordinate system
water_travis <- st_transform(water_travis, crs = st_crs(tracts_travis))

# Erase water from tracts using st_difference
tracts_nowater <- st_difference(tracts_travis, st_union(water_travis))

```

#export 
```{r}


```



#BONUS WORK


Create code that cleans up bad variables
```{r}
# Drop geometry before processing
tract_data_clean <- tract_data %>% st_drop_geometry()

# === Remove Variables with High Relative MOE ===
#create a named vector'rel_moe, which has names of variables and a calculated relative margin of error (percentage of all rows that have relative margins of error >50% of the estimate)
rel_moe <- sapply(grep("E$", names(tract_data_clean), value = TRUE), function(est_col) {
  moe_col <- sub("E$", "M", est_col)
  if (moe_col %in% names(tract_data_clean)) {
    rel <- tract_data_clean[[moe_col]] / tract_data_clean[[est_col]]#calculates the value of error/estimate into a new column
    return(mean(rel > 0.5, na.rm = TRUE)) # in the new column , how many values in a variable have a high relative margin of error. 
    
    #this is what create the rel_moe named vecotor - variable to percent of rows with rel >.5. this is used to select remaining variables.


  }
  return(NA)
})


vars_to_remove <- names(rel_moe[rel_moe > 0.5]) #uses named vector to say, if value is over .5, remove
moe_to_remove <- sub("E$", "M", vars_to_remove) #subtract columns of those variables

# Map renamed fields if they exist
renamed_vars <- names(tract_data_final)
]

print("Variables removed due to high relative MOE (>50%):")
print(readable_removed)

```
