---
title: "FBC_Health"
output: html_document
date: "2023-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(sf)
library(osmdata)
library(tigris)

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

crs <- "EPSG:2248"


# set bounding box - the maximum x-y extent you are interested in
# Do some right-clicking in google maps to figure this out



q0 <- opq(bbox = c(-77.007980,38.651735,-76.565094,39.082639)) 

park <- add_osm_feature(opq = q0, key = 'leisure', value = "park") %>%
  osmdata_sf(.)


park.sf <- st_geometry(park$osm_polygons) %>%
  st_transform(4326) %>%
  st_sf() %>%
  cbind(., park$osm_polygons$name) %>%
  rename(NAME = park.osm_polygons.name)

restaurant <- add_osm_feature(opq = q0, key = 'amenity', value = "restaurant") %>%
  osmdata_sf(.)

restaurant.sf <- st_geometry(restaurant$osm_points) %>%
  st_transform(4326) %>%
  st_sf() %>%
  cbind(., restaurant$osm_points$name) %>%
  cbind(., restaurant$osm_points$osm_id) %>%
  cbind(., restaurant$osm_points$addr.street) %>%
  cbind(., restaurant$osm_points$addr.postcode) %>%
  cbind(., restaurant$osm_points$addr.city) %>%
  cbind(., restaurant$osm_points$addr.state) %>%
  cbind(., restaurant$osm_points$cuisine)

# Remove "restaurant.osm_points." string from column names
colnames(restaurant.sf) <- gsub("restaurant.osm_points.","",colnames(restaurant.sf))

# Add the features for Philadelphia County
# You can use this to clip things if it's necessary

paCounties <- counties(state = 'PA', cb = FALSE)

phila <- paCounties %>%
  filter(NAME == "Philadelphia") %>%
  st_as_sf() %>%
  st_transform(4326)

# Map it

ggplot()+
  geom_sf(data = phila, color = "black")+
  scale_color_manual(values = c("orange", "blue"))+
  geom_sf(data = park.sf, fill = "light green", color = "transparent")
