---
title: "processing_pdfs_orderedcombined"
author: "Saturnya"
date: "2025-04-24"
output: html_document
---

This R Markdown document demonstrates how to combine multiple PDF files into a single PDF using the **pdftools** package, wrapped in a reusable function that supports an overwrite feature.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Install and Load **pdftools**

```{r install-pdftools}
if (!requireNamespace("pdftools", quietly = TRUE)) {
  install.packages("pdftools")
}
library(pdftools)
```

## 2. Define the `combine_pdfs()` Wrapper Function

The function takes three arguments:

- `input_files`: character vector of PDF file paths, in the order to combine
- `output_file`: path for the combined PDF
- `overwrite`: logical, whether to overwrite an existing output_file (default `FALSE`)

```{r define-function}
combine_pdfs <- function(input_files, output_file, overwrite = FALSE) {
  # Check overwrite
  if (file.exists(output_file) && !overwrite) {
    stop("Output file already exists. Use overwrite = TRUE to replace it.")
  }

  # Ensure output directory exists
  out_dir <- dirname(output_file)
  if (!dir.exists(out_dir)) {
    dir.create(out_dir, recursive = TRUE)
  }

  # Attempt to combine
  tryCatch(
    {
      pdf_combine(input = input_files, output = output_file)
      message(sprintf(
        "✅ Successfully combined %d PDFs into '%s'",
        length(input_files), output_file
      ))
    },
    warning = function(w) {
      message("⚠️ Warning during combine: ", w$message)
    },
    error = function(e) {
      stop("❌ Failed to combine PDFs: ", e$message)
    }
  )
}
```

## 3. Specify and Combine Your PDFs

Edit the input vector and output path, then call the function. If you want to overwrite an existing file, set `overwrite = TRUE`.

```{r example-use}
pdf_files <- c(
  "C:/Users/User/Downloads/LETTER_ANDREASGAZEL_GE_04232025.pdf",
  "C:/Users/User/Downloads/CA_GE_04242025_v1.pdf"
)
output_pdf <- "C:/Users/User/Desktop/PROPOSAL_GE_04242025.pdf"

# Combine, allowing overwrite if needed:
combine_pdfs(pdf_files, output_pdf, overwrite = TRUE)
```

---

**Customization Ideas**

- Wrap in an R package function for reuse across projects.
- Add interactive prompts to select files at runtime.
- Incorporate logging or verbose modes for detailed progress.
- Extend to batch process directories of PDFs.


```

