---
title: "Final Study Area Map Rendering"
author: "Saturnya"
date: "2025-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
source(here("Documents","TOOLKIT","R","utils", "functions.R"))
library(classInt)
library(scales)
library(MetBrewer)


```
#PRESETS

1. CRS- EPSG:2277
2. COLUMN NAME ORDER FOR THE DF


#Introduction

The following code laods objects that will be layered/processed to become final maps for the client.

The plotted map will include: a basemap, cbsa statistical boundary for austin, census tract boundaries without water layer,primary roads, client's home site, 

#Load Objects

##census tracts data for austin's county with viable variables houtput
```{r}
final_tracts <- readRDS(
  file = here("clients", "arshad", "cache", "austin_acs5yr_tracts_df_houtput.rds")
)

```


###hotspots per variables
Convert the og object into long format for the ability to identify hotspot tracts.

```{r}

hotspot_long <- final_tracts %>%
  select(geometry, NAME, ends_with("_hotspot")) %>%
  pivot_longer(
    cols = -c(geometry, NAME),
    names_to = "variable",
    values_to = "hotspot"
  ) %>%
  mutate(#this part creates the name of columns for faceted/iterated ploting
    variable = gsub("_hotspot", "", variable),
    hotspot_status = case_when(
      is.na(hotspot) ~ "Unavailable",
      hotspot ~ "Hotspot",
      TRUE ~ "Not Hotspot"
    ),
    hotspot_status = factor(
    hotspot_status,
    levels = c("Hotspot", "Not Hotspot", "Unavailable"))
  ) %>%
  st_as_sf()

##LOOK INTO THE NECCESITY OF THIS CODE BELOW, CAN IT BE INCLUDED IN THE FUNCTION ABOVE?
#to ensure it works, no 0 length variables
hotspot_long_clean <- hotspot_long %>%
  group_by(variable) %>%
  filter(!all(is.na(hotspot_status))) %>%
  filter(n_distinct(hotspot_status, na.rm = TRUE) > 1) %>%
  ungroup()

```

## Roads

Load from local data

create a major highways object

```{r}
# Step 1: Define zip path
roads_zip <- "C:/Users/User/Desktop/local data/tl_2024_texas_prisecroads.zip"

# Step 2: Extract files
unzipped_dir <- extract_zip_with_fallback(roads_zip)

# Step 3: Find and load shapefile
#shp_file <- list.files(unzipped_dir, pattern = "\\.shp$", full.names = TRUE)[1]
roads <- st_read(unzipped_dir)

# Step 4: Filter Interstates and U.S. Highways
major_highways <- roads %>%
  filter(RTTYP %in% c("I", "U"))

# Step 5: Reproject to EPSG:2277 (Texas Central)
major_highways_2277 <- st_transform(major_highways, crs = 2277)

```

##CBSA
From local data folder off all us cbsa, filter the geometry of austin.
```{r}

# Step 1: Define path
cbsa_zip <- "C:/Users/User/Desktop/local data/tl_2024_us_cbsa.zip"

# Step 2: Extract
unzipped_cbsa <- extract_zip_with_fallback(cbsa_zip)

# Step 3: Locate shapefile and read
cbsa_shp <- list.files(unzipped_cbsa, pattern = "\\.shp$", full.names = TRUE)[1]
cbsa_all <- st_read(cbsa_shp)

# Step 4: Filter to Austin CBSA
austin_cbsa <- cbsa_all %>%
  filter(NAMELSAD == "Austin-Round Rock-San Marcos, TX Metro Area")

# Step 5: Reproject to EPSG:2277
austin_cbsa_2277 <- st_transform(austin_cbsa, crs = 2277)
```
##Water
```{r}
travis_water_2277 <- extract_zip_with_fallback("C:/Users/User/Desktop/local data/tl_2024_48453_areawater.zip") %>%
  st_transform(crs = 2277)

travis_water_2277 <- travis_water_2277 %>%
  filter(AWATER > 1e5) 
```


##Site Point

```{r}
#crs already confirmed
AS_home<-readRDS(file = here("clients", "arshad", "cache", "arshad_austin_home_site.rds"))


```
## Austin city boundary

```{r}





```


#Final plots
The following will be pdf outputs that take in as arguments each of the obejcts loaded above. 
##Chloropleths

Get the input df for this from the 

Cut off the hotspot variables from the dataset (keep the n_hotspots)
```{r}
#Get these from the expsum refined tract data output PLUS
##Include counts of POI's per tract

generate_choropleth_maps(
  geodata = final_tracts,
  site_points = home_site,
  water = travis_water_2277,
  roads = major_highways_2277,
  cbsa_boundary = austin_cbsa_2277,
  overwrite = TRUE,          # to allow regenerating output files
  run_leaflet = FALSE        # turn off Leaflet if it throws errors
)
```

##Hotspots
###custom export function for hotspots

```{r}

generate_faceted_hotspot_maps(
  hotspot_data = hotspot_long_clean,       # your long-format hotspot sf object
  site_points = home_site,                 # the green home location point
  water = travis_water_2277,               # water body sf
  roads = major_highways_2277,             # roads shapefile already projected
  cbsa_boundary = austin_cbsa_2277,        # CBSA area to clip context layers
  overwrite = TRUE                         # set TRUE to regenerate outputs
)



```
###FUNCTION TO COMBINE THOSE MAPS SO FAR

```{r}
combine_choropleth_and_hotspot_pdfs()

```



##COMPOSITE SCORE MAP- ANNOTATION
```{r}
#We have the tmap at the end of the locmor_plots script


```

##Distance accumulation from the house
```{r}


```

#COMPOSITE SCORE CODE
```{r}

#I MADE THE COMPOSITE OBJECT USING A PYTHON SCRIPT IN PyTOOLKIT - saved as geopackage and am loading now

## 2.1  Composite layer with tier column
comp_path <- "C:/Users/User/Documents/TOOLKIT/clients/arshad/cache/arshad_austin_acs_final_stackscore.gpkg"
composite_sf <- st_read(comp_path, quiet = TRUE) |>
  st_transform(2277)

## 2.2  Roads (folder already unzipped)
roads_dir <- "C:/Users/User/Desktop/local data/texas_roads"
roads_shp <- list.files(roads_dir, pattern = "\\.shp$", full.names = TRUE)[1]
roads     <- st_read(roads_shp, quiet = TRUE) |>
             st_transform(2277)

major_highways_2277 <- roads |>
  filter(RTTYP %in% c("I", "U"))          # Interstates & U.S. highways

## 2.3  Water (folder already unzipped)
water_dir <- "C:/Users/User/Desktop/local data/texas_water"
water_shp <- list.files(water_dir, pattern = "\\.shp$", full.names = TRUE)[1]
travis_water_2277 <- st_read(water_shp, quiet = TRUE) |>
                     st_transform(2277) |>
                     filter(AWATER > 1e5)            # keep larger bodies

## 2.4  CBSA boundaries (folder already unzipped)
cbsa_dir <- "C:/Users/User/Desktop/local data/us_cbsa"
cbsa_shp <- list.files(cbsa_dir, pattern = "\\.shp$", full.names = TRUE)[1]
austin_cbsa_2277 <- st_read(cbsa_shp, quiet = TRUE) |>
  filter(NAMELSAD == "Austin-Round Rock-San Marcos, TX Metro Area") |>
  st_transform(2277)

## 2.5  Home‑site point
site <- readRDS("C:\\Users\\User\\Documents\\TOOLKIT\\clients\\arshad\\cache\\arshad_austin_home_site.rds") |>
        st_transform(2277)

# ------------------------------------------------------------
# 3.  Colour palette for 5 tiers
# ------------------------------------------------------------
library(MetBrewer)

# 1.  Get the *actual* tier values that exist in the data
lvls <- sort(unique(composite_sf$tier))   # character vector, e.g. c("1","2","3","4","5","Other")

# 2.  Build a palette that keeps Signac for main tiers, custom for Other
n_main  <- sum(lvls != "Other")           # how many non‑Other tiers
pal_vec <- MetBrewer::met.brewer("Signac", n_main)
names(pal_vec) <- lvls[lvls != "Other"]   # exact match

pal_vec["Other"] <- "#dfffcf"             # pastel green
pal_vec <- pal_vec[lvls]                  # re‑order to data order

# 3.  Call the function
generate_composite_tier_map(
  composite_sf   = composite_sf,
  tier_col       = "tier",
  water          = travis_water_2277,
  roads          = major_highways_2277,
  site_points    = site,
  cbsa_boundary  = austin_cbsa_2277,
  tier_palette   = pal_vec,
  filename       = "arshad_composite_tier_map_large.pdf",
  dpi            = 300,
  overwrite      = TRUE
)
```
```{r}





```

