---
title: "Local Moran's I -Hotspot analysis"
author: "Saturnya"
date: "2025-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

crs <- "EPSG:2277"
#install.packages("spdep")
#install.packages("tidyr")
#install.packages("ggplot2")
install.packages("tmap")
library(sf)
library(spdep)
library(tmap)
library(here)
library(ggplot2)
library(tidyr)
library(dplyr)
library(stringr)

source(here("R","utils", "functions.R"))

```
#Local Moran's I Hotspot Analysis

##Load project-level cleaned df with points of interests cached dataset.

This is a dataset that only has tract level variables that we are interest in for this project

The object is tract level ACS 5-Year estimates with counts of point of interst location per tract.

```{r}

#Cache
locmor_working_df<-readRDS(here("clients", "arshad", "cache", "arshad_austin_tracts_ready_poi.rds"))

#for 

locmor_working_df <- locmor_working_df %>%
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

```

##Create neighbors matrix 

```{r}
neighbors <- poly2nb(locmor_working_df)

```

##Select numeric variables in dataset - (precheck)

Verify count of NA values. I did, ready to go removing NA values when iterating over each variable.

```{r}
variables <- names(locmor_working_df)[sapply(locmor_working_df, is.numeric)]

#How many NA values per variables?
##Initialize summary table for post calculations
summary_df <- data.frame(variable = character(),
                         n_NA = integer(),
                         n_hotspots = integer(),
                         stringsAsFactors = FALSE)




```
##Function to iterate local moran's I calculations on a df object

The following code iterates over each variable, removes NA values, subsets the neighbors and weights, then calculates the local morans values and identifies hotspot.

Note that you can change the threshhold of the p value. .05 is good, 0.03 even better. 
```{r}

for (var in variables) {
  if (all(is.na(locmor_working_df[[var]]))) next

  subset_logical <- !is.na(locmor_working_df[[var]])
  non_na_index <- which(subset_logical)

  subset_neighbors <- subset.nb(neighbors, subset = subset_logical)
  subset_weights <- nb2listw(subset_neighbors, style = "W", zero.policy = TRUE)
  subset_values <- locmor_working_df[[var]][non_na_index]

  local_moran <- localmoran(subset_values, subset_weights, zero.policy = TRUE)

  I_vals <- rep(NA, nrow(locmor_working_df))
  p_vals <- rep(NA, nrow(locmor_working_df))
  hotspots <- rep(NA, nrow(locmor_working_df))

  I_vals[non_na_index] <- local_moran[, 1]
  p_vals[non_na_index] <- local_moran[, 5]
  hotspots[non_na_index] <- p_vals[non_na_index] < 0.03 & I_vals[non_na_index] > 0

  locmor_working_df[[paste0(var, "_I")]] <- I_vals
  locmor_working_df[[paste0(var, "_p")]] <- p_vals
  locmor_working_df[[paste0(var, "_hotspot")]] <- hotspots

  summary_df <- rbind(summary_df, data.frame(
    variable = var,
    n_NA = sum(is.na(locmor_working_df[[var]])),
    n_hotspots = sum(hotspots, na.rm = TRUE)
  ))
}


# the result is a summary df and more columns in the df for the individual local morans i outputs
```

###Clean up the working dataset


```{r}
# Remove columns ending in _p or _I
locmor_working_df <- locmor_working_df %>%
  select(-matches("(_p|_I)$"))

# Remove columns with more than 40% missing values
threshold_1 <- 0.4

locmor_working_df <- locmor_working_df %>%
  select(where(~ mean(is.na(.)) <= threshold_1))
```



#Calculate times that a variable is a hotspot in an area

##First, Load all the variables that matter. 

###Load column names and select only the variables that you want to take into consideration

The code below makes our wide data into long with the pivot function.

because I want to count the times a variable is a hot spot, and want to display it later, this code strips the 'hotspot part of the variable.

make sure factor columns are exactly that. as oppsoed to binary. 

###The code below takes the output local morans df and pivots it to calculate counts of hotspots

Pivot longer makes the data pivot so that per variable, you have a true or false value per tract.

###keep the geometry and columns that end with _hotspot
These columns maintain the TRUE/FALSE value of if a particular variable was a hot spot.

```{r}
##KEEP THIS NAMED OBJECT IN MIND IS THE LOCAL MORANS OUTPUTS DF
###THIS CODE CALCULATES WHICH WERE HOTSPOTS AND RENAMES VARIABLES TO BE ABLE TO PLOT CLEANLY
# Stack all hotspot columns into long format

hotspot_long <- locmor_working_df %>%
  select(geometry, NAME, ends_with("_hotspot")) %>%
  pivot_longer(
    cols = -c(geometry, NAME),
    names_to = "variable",
    values_to = "hotspot"
  ) %>%
  mutate(#this part creates the name of columns for faceted/iterated ploting
    variable = gsub("_hotspot", "", variable),
    hotspot_status = case_when(
      is.na(hotspot) ~ "Unavailable",
      hotspot ~ "Hotspot",
      TRUE ~ "Not Hotspot"
    ),
    hotspot_status = factor(
    hotspot_status,
    levels = c("Hotspot", "Not Hotspot", "Unavailable"))
  ) %>%
  st_as_sf()

##LOOK INTO THE NECCESITY OF THIS CODE BELOW, CAN IT BE INCLUDED IN THE FUNCTION ABOVE?
#to ensure it works, no 0 length variables
hotspot_long_clean <- hotspot_long %>%
  group_by(variable) %>%
  filter(!all(is.na(hotspot_status))) %>%
  filter(n_distinct(hotspot_status, na.rm = TRUE) > 1) %>%
  ungroup()

```

#Export the local morans I object


###Consider arranging the maps so that specific variabels are next to eachother
Use this chunk to arrange the facetted maps in a best way for side by side views. 

```{r}


#put variables that you can put side by side here

```

###Save the Maps as Images and PDF w/ 'ggforce' and custom function This creates pdf of the distribution of identified hotspots per variable (all) **manually select for client report.

This is where some extra customization can occur.

YOu can choose font families.

gg base is a plot with set parameters that then is copied in style for each variable
```{r}
#master function to produce images and pdfs from a ggplot object

export_faceted_map_pages <- function(data, gg_base, facet_var, file_basename, ncol = 2, nrow = 1,
                                     img_dir = here("clients", "arshad", "outputs", "img"),
                                     pdf_dir = here("clients", "arshad", "outputs", "pdf"),
                                     width = 14, height = 8, dpi = 300, overwrite = TRUE) {
  
  # Create output dirs if needed
  if (!dir.exists(img_dir)) dir.create(img_dir, recursive = TRUE)
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)

  # Calculate number of pages
  n_panels <- length(unique(data[[facet_var]]))
  n_pages <- ceiling(n_panels / (ncol * nrow))

  # Define output paths
  pdf_path <- file.path(pdf_dir, paste0(file_basename, ".pdf"))
  png_paths <- file.path(img_dir, paste0(file_basename, "_page_", seq_len(n_pages), ".png"))

  # Check for existing files if overwrite = FALSE
  if (!overwrite) {
    existing_files <- c(pdf_path, png_paths[file.exists(png_paths)])
    if (file.exists(pdf_path) || length(existing_files) > 0) {
      stop("🛑 Some output files already exist. Set `overwrite = TRUE` to replace them.")
    }
  }

  # Remove old files if overwrite = TRUE
  if (overwrite) {
    if (file.exists(pdf_path)) {
      file.remove(pdf_path)
      message("⚠️ Existing PDF removed: ", pdf_path)
    }
    for (png_path in png_paths) {
      if (file.exists(png_path)) {
        file.remove(png_path)
        message("⚠️ Existing PNG removed: ", png_path)
      }
    }
  }

  # Open PDF device
  pdf(pdf_path, width = width, height = height)

  # Loop through and export
  for (i in seq_len(n_pages)) {
    p <- gg_base +
      ggforce::facet_wrap_paginate(as.formula(paste0("~", facet_var)),
                                   ncol = ncol, nrow = nrow, page = i)

    # Save PNG
    ggsave(png_paths[i], plot = p, width = width, height = height, dpi = dpi)

    # Add to PDF
    print(p)
  }

  dev.off()
  message("✅ Export complete: ", n_pages, " page(s) written.")
}
```
#### Use Base Object

Create a base object prior to faceting

considering editing this specific base object!!!! as a function in itself
```{r}
hotspot_base <- ggplot(hotspot_long_clean) +
  geom_sf(aes(fill = hotspot_status), color = "grey90", size = 0.1) +
  scale_fill_manual(
    values = c("Hotspot" = "red", "Not Hotspot" = "white", "Unavailable" = "lightgrey"),
    na.value = "grey80", drop = FALSE
  ) +
  labs(fill = "Hotspot Status") +
  theme_void(base_size = 14) +
  theme(
    strip.text = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(10, 10, 10, 10)
  )



```
#### Call the custom function to export pdfs of hotspotclean
Consider how to use basename part of function to get it where it needs to go.


Keep in mind: the internal function must be changed client to client.
```{r}
#This function saves the faceted maps for the variable in the correct client folders.
export_faceted_map_pages(
  data = hotspot_long_clean,
  gg_base = hotspot_base,
  facet_var = "variable",
  file_basename = "hotspot_facets"
)

#success

```
SUCCESS!! 

#Next,identify the tract the has the most amount of hotspots. this creates a percensus tract count of hotspots

Filter from hot spot long the variables that interest you.

Remove redundant variables that may cause a double
##remove unwanted variables (for the count)
This is where you filter out the varibales you do not want use for hot spot count
```{r}
#remove from hotspot_long variables that are not desired for per census tract score.
#they are in rows, filter the variable column for what you want
hotspot_filtered <- hotspot_long_clean %>%
  filter(variable %in% c(
    "total", "spa", "aesthetics", "income_125k_plus.y",
    "male_30_54", "female_30_54", "edu_attain_pct",
    "hispanic_or_latino_pct", "white_alone_pct", "aggregate_household_income_pct",
    "aggregate_household_income_clipped", "median_household_income_clipped",
    "log_median_household_income", "log_aggregate_household_income"
  ))

#Take this OPPURTUNITY TO REMOVE VARIABLESFROM THE COUNT
```

Now, get the count, per census tract ID where there is a value of True. 
## perform the summary count, and object to be mapped
Create a histogram of values. 

******************consider which variables actually count ***************************

```{r}
hotspot_counts <- hotspot_filtered %>%
  group_by(NAME) %>%
  summarise(n_hotspots = sum(hotspot, na.rm = TRUE)) %>% #counts the number of True by default (1 or 0)
  ungroup() #ungroup to end the grouping structure so that future operations arent grouped
```
##Look at Histogram 'hotspot count' is the summary object that contains the per tract information on hotspot counts

Use histogram to detemine cutoffs for 'hottest' census tract
```{r}
#plot historgram


ggplot(hotspot_counts, aes(x = n_hotspots)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  geom_text(stat = "bin", binwidth = 1, aes(label = ..count..), 
            vjust = -0.5, color = "black", size = 4) +
  labs(
    title = "Distribution of Hotspots per Census Tract",
    x = "Number of Hotspots",
    y = "Count of Census Tracts"
  ) +
  theme_minimal()
```

## Visualize hottest tracts

Filter hottest tracts and add a label to the initial dataframe 'locmor_austin_df' that includes them. Save as a new object.

Note: in dplyr joins only one object should be sppatial

###Filter for hottest based on histogram 'hottest' object

This is where you enter the 'hottest spots'

For 'hot', a census tract had to be a hot spot at least 3 times


```{r}


#filter for hottest tracts, visualize on a map

austin_hottracts <- hotspot_counts %>%
  filter(n_hotspots > 5) %>%
  select(-geometry)%>%
  mutate(hottest = case_when(
         n_hotspots > 7 ~ "hottest", TRUE ~ "hot")
 )

```


###now I will add these labels back to the original dataframe before 

dyplr can only join one spatial object
```{r}


austin_acs5yr_tracts_df_houtput <- locmor_working_df %>%
  left_join(
    austin_hottracts %>%
      st_drop_geometry() %>%
      select(NAME, hottest),
    by = "NAME"
  ) %>%
  mutate(hottest = replace_na(hottest, "not hot"))

```

###Also, add the number of hotspots to the loc_mor output

```{r}
austin_acs5yr_tracts_df_houtput<-austin_acs5yr_tracts_df_houtput%>%
   left_join(hotspot_counts%>%
               st_drop_geometry()
               , by="NAME")

```
###Save the df of hottest

```{r}
saveRDS(austin_acs5yr_tracts_df_houtput, here("clients", "arshad", "cache", "austin_acs5yr_tracts_df_houtput.rds"))

```
##Visualize it hottest and hot tracts

```{r}
# Set map mode
tmap_mode("plot")  # or "view" for interactive

# Plot with layered styling
arshad_austin_locmor_hottest_tmap<-tm_shape(austin_acs5yr_tracts_df_houtput) +
  # Base layer: all tracts in grey
  tm_polygons(col = "lightgrey", border.col = "white") +
  
  # Overlay: only hot/hottest tracts
  tm_shape(austin_acs5yr_tracts_df_houtput %>% filter(hottest != "not hot")) +
  tm_polygons("hottest", 
              palette = c("hot" = "orange", "hottest" = "red"), 
              title = "Hotspot Level") +
  
  # Optional: add tract NAME labels to hot/hottest
  #tm_text("NAME", size = 0.5, auto.placement = TRUE) +
  
  # Layout styling
  tm_layout(main.title = "Austin Tracts: Hot & Hottest",
            legend.outside = TRUE)

#Cache
saveRDS(arshad_austin_locmor_hottest_tmap, here("clients", "arshad", "cache", "arshad_austin_locmor_hottest_tmap.rds"))



```

###SAVE as image and pdf

Function to check if file exists and save both png and pdf.

#### Custom function that saves both a pdf and img of a map 
The function contains and overwrite feature.

Keep in mind the width and lenght.

You will have to save these to the right places per client
```{r}
#KEEP IN MIND THE IMG AND PDF HAVE TO BE DIRECTED TO the right client folder
export_tmap_plot <- function(tmap_plot, file_basename,
                             img_dir = here("clients", "arshad", "outputs", "img"),
                             pdf_dir = here("clients", "arshad", "outputs", "pdf"),
                             width = 14, height = 8, dpi = 300, units = "in",
                             overwrite = TRUE) {
  
  # Create output dirs if they don't exist
  if (!dir.exists(img_dir)) dir.create(img_dir, recursive = TRUE)
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)

  # Define output paths
  png_path <- file.path(img_dir, paste0(file_basename, ".png"))
  pdf_path <- file.path(pdf_dir, paste0(file_basename, ".pdf"))

  # Handle overwriting
  if (overwrite) {
    if (file.exists(png_path)) {
      file.remove(png_path)
      message("⚠️ Existing PNG removed: ", png_path)
    }
    if (file.exists(pdf_path)) {
      file.remove(pdf_path)
      message("⚠️ Existing PDF removed: ", pdf_path)
    }
  } else {
    if (file.exists(png_path) || file.exists(pdf_path)) {
      stop("🛑 File already exists. Set `overwrite = TRUE` to replace it.")
    }
  }

  # 🔧 Save the plot
  tmap::tmap_save(tmap_plot, filename = png_path, width = width, height = height, dpi = dpi, units = units)
  tmap::tmap_save(tmap_plot, filename = pdf_path, width = width, height = height, units = units)

  message("✅ tmap plot saved as PNG and PDF.")
}
```

#### Save the Hottest Spots plot with function above

Transfer the function to function.r file



```{r}
export_tmap_plot(
  tmap_plot = arshad_austin_locmor_hottest_tmap,
  file_basename = "hottest_locmor_tmap_plot",
  overwrite = TRUE
)

```
