---
title: "Merge: NY State-Owned Parcels / NYC MAPPLUTO As CSV"
author: "SaturnyaLLC"
date: "2025-03-14"
output: html_document
---

```{r setup, include=FALSE}  
knitr::opts_chunk$set(echo = TRUE, messages=FALSE)
```
#Load neccesary packages.

```{r}
#load packages

install.packages("sf")
install.packages("dplyr")
install.packages("tidyverse")
install.packages("readr")
library(sf)
library(dplyr)
library(tidyverse)
library(readr)

```


#Introduction
The purpose of this script is to import NY State-Owned Parcels and filter them by borough.

#Data import
Make sure that the CRS is the same before performing any spatial operations. 
```{r}

#open online NY STATE OWNED PARCELS geojson for reduction

nystateownedparcels<-st_read("https://data.gis.ny.gov/datasets/sharegisny::nys-tax-parcels-state-owned.geojson")

#open NYC MAPPLUTO file (from local machine)

# Read the zipped File Geodatabase directly. The previous produced errors, there is a way to load in specifiv variable contained in the dataset. The gollowing code allows you to see a list of layers available for reading. You use the title of the file as an argument in the st_read(). 

nyc_mappluto <- st_read("C:/Users/User/Desktop/local data/MapPLUTO24v4_1.gdb", layer = "MapPLUTO_24v4_1_clipped")


#CRS WORK
#first, get CRS of object that you want to analyze. In this case, the mappluto crs.

crs_reference <- st_crs(nyc_mappluto)

# Set the CRS of the first object (target object) to match the CRS of the reference object. st_transforms performs this function.
nycwide_stateowned_parcels <- st_transform(nycwide_stateowned_parcels, crs = crs_reference)

```

#Data Wrangling

##NY STATE DATA

The following code filters the NY STATE OWNED PARCELS data set by borough and vacant land PROP CLASS values. In NYC, Vacant land is given a PROP_CLASS value of 11

To create a clean excel for organizers, I dropped the geometry before saving.

###Filtering for NYC

```{r}

bronx_stateowned_parcels<-nystateownedparcels %>%
    filter(
      grepl("bronx", MUNI_NAME, ignore.case = TRUE)&
      grepl("11", PROP_CLASS)
    )%>%
     select(COUNTY_NAME, MUNI_NAME, PARCEL_ADDR, LOC_ST_NBR, LOC_STREET,               LOC_ZIP,PROP_CLASS, LAND_AV, TOTAL_AV, SQ_FT, BLDG_STYLE, 
            BLDG_STYLE_DESC, PRIMARY_OWNER, NYS_NAME
     )
      

manhattan_stateowned_parcels<-nystateownedparcels %>%
    filter(
      grepl("manhattan",MUNI_NAME, ignore.case = TRUE)&
      grepl("11", PROP_CLASS)
    )%>%
     select(COUNTY_NAME, MUNI_NAME, PARCEL_ADDR, LOC_ST_NBR, LOC_STREET,               LOC_ZIP,PROP_CLASS, LAND_AV, TOTAL_AV, SQ_FT, BLDG_STYLE, 
            BLDG_STYLE_DESC, PRIMARY_OWNER, NYS_NAME
     )
      

brooklyn_stateowned_parcels<-nystateownedparcels %>%
    filter(
      grepl("brooklyn",MUNI_NAME, ignore.case = TRUE)&
      grepl("11", PROP_CLASS)
    )%>%
     select(COUNTY_NAME, MUNI_NAME, PARCEL_ADDR, LOC_ST_NBR, LOC_STREET,               LOC_ZIP,PROP_CLASS, LAND_AV, TOTAL_AV, SQ_FT, BLDG_STYLE, 
            BLDG_STYLE_DESC, PRIMARY_OWNER, NYS_NAME
     )

queens_stateowned_parcels<-nystateownedparcels %>%
    filter(
      grepl("queens",MUNI_NAME, ignore.case = TRUE)&
      grepl("11", PROP_CLASS)
    )%>%
     select(COUNTY_NAME, MUNI_NAME, PARCEL_ADDR, LOC_ST_NBR, LOC_STREET,               LOC_ZIP,PROP_CLASS, LAND_AV, TOTAL_AV, SQ_FT, BLDG_STYLE, 
            BLDG_STYLE_DESC, PRIMARY_OWNER, NYS_NAME
     )

staten_stateowned_parcels<-nystateownedparcels %>%
    filter(
      grepl("staten island",MUNI_NAME, ignore.case = TRUE)&
      grepl("11", PROP_CLASS)
    )%>%
     select(COUNTY_NAME, MUNI_NAME, PARCEL_ADDR, LOC_ST_NBR, LOC_STREET,               LOC_ZIP,PROP_CLASS, LAND_AV, TOTAL_AV, SQ_FT, BLDG_STYLE, 
            BLDG_STYLE_DESC, PRIMARY_OWNER, NYS_NAME
     )
```

### Combine the individual boroughs into 1 dataset
The following code takes the individual datasets and combines their rows to create a city-wide dataset

```{r}
#creates an object containing all rows for NYC

nycwide_stateowned_parcels<-rbind(bronx_stateowned_parcels,                   manhattan_stateowned_parcels, queens_stateowned_parcels,                  staten_stateowned_parcels, brooklyn_stateowned_parcels)


```

## Join the State data to the NYC MAPPLUTO Data to get Block-Lot-Combos
The following code intersects NY State owned parcels with NYC mappluto to attempt to join addresses and block-lot combos with parcels

```{r}
nyc_stateowned_mappluto<-st_join(nycwide_stateowned_parcels, nyc_mappluto)

nyc_stateowned_mappluto <- nyc_stateowned_mappluto %>%
  st_drop_geometry()%>%
  distinct(Address, .keep_all = TRUE)%>%
  filter(grepl("^[XOC]$", OwnerType, ignore.case = TRUE))%>%
  select(PROP_CLASS,LAND_AV,TOTAL_AV,SQ_FT,BLDG_STYLE,PRIMARY_OWNER,Borough,Block,Lot,CD,Council,ZipCode,PolicePrct,Address,ZoneDist1,ZoneDist2,ZoneDist3,ZoneDist4,Overlay1,Overlay2,SPDist1,SPDist2,SPDist3,OwnerType,OwnerName,LotArea,BldgArea,AssessLand,AssessTot,BuiltFAR,ResidFAR,CommFAR,FacilFAR,BBL,TaxMap,ZoneMap)

#use this code to arrange columns as desired by their position. Use colnames(df) to identify the indexed position of the column.

nyc_stateowned_mappluto <- nyc_stateowned_mappluto %>%
  select(
    !!colnames(nyc_stateowned_mappluto)[34],  # 34th column
    !!colnames(nyc_stateowned_mappluto)[14],  # 14th column
    !!colnames(nyc_stateowned_mappluto)[25],  # 25th column
    !!colnames(nyc_stateowned_mappluto)[6],   # 6th column
    everything()  # All remaining columns
  )

#the preceding object does not have geometry!

```

#Data Export
The following code first creates the desired path. It then checks if the file already exists to avoid duplicating it. Lastly, it saves the file in the desired format.

```{r}
#desired path establishment

bronx_file_path <- "C:/Users/User/Documents/NEHEMIAH/Outputs/datasets/nehemiah_bronx_stateowned_parcels_03142025.csv"

#function that checks its existence then saves. !file.exist relates the TRUE/FALSE result to the 'if' function, telling it to run or not if the file exists. Make sure the the object you want to save is named correctly


# Check if the file exists
if (!file.exists(bronx_file_path)) {
  # Example dataset: Replace this with your actual data
  df <- bronx_stateowned_parcels  # Assuming you already have the filtered dataset
  
  # Save as CSV if the file doesn't exist
  write_csv(df, bronx_file_path)
  
  message("File did not exist. Saved: ", bronx_file_path)
} else {
  message("File already exists. No action taken.")
}

# do this for EACH OBJECT TO SAVE INTO THE CLIENTS OUTPUT FOLDER!

nyc_stateowned_mappluto_filepath<-"C:/Users/User/Documents/NEHEMIAH/Outputs/datasets/nehemiah_nycwide_stateowned_mappluto_03142025.csv"

  if (!file.exists(nyc_stateowned_mappluto_filepath)) {
  # Example dataset: Replace this with your actual data
  df <- nyc_stateowned_mappluto  # Assuming you already have the filtered dataset
  
  # Save as CSV if the file doesn't exist
  write_csv(df, nyc_stateowned_mappluto_filepath)
  
  message("File did not exist. Saved: ",nyc_stateowned_mappluto_filepath)
} else {
  message("File already exists. No action taken.")
}
  
```