---
title: 'OSM: Points of Interest by client'
author: "Saturnya"
date: "2025-04-10"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(tigris_use_cache = TRUE)




install.packages(c("tigris","purrr","osmdata"))
library(tigris)
library(sf)
library(dplyr)
library(purrr)
library(tmap)
library(osmdata)

```

## R Markdown

```{r}

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```
#Introduction

The following document will return an sf object of points for a given study area and a particular use. The example here in is for medical spas.

In this code, the study area will be named, the client will be named, and file names will be generated as a function of these inputs.

I will use an open street maps query to get the location of medical spas in austin texas as points. The return is a json object that must be processed for proper visualization. 


#OSM Process

##Load Study area from open street maps

This will represent the boudns of the study area. Does OSM keep CBSA Boundaries? Should be quick to load.

```{r}



```




##create query

Ensure that it stays with the bounds of the study area

```{r}



```


##Extract Data

Process the JSON returned 

```{r}



```

##Filter

```{r}



```



##Extract points
```{r}



```


##convert to Sf

```{r}


```




#export sf

AS RDS object. To be loaded elsewhere as a layer for a plot. 
```{r}



```

